---
title: "preparing_ras_stack"
output: html_document
date: "2024-11-21"
---
# libraries
```{r}
options(java.parameters = "-Xmx24G" )
library (sf) 
library(terra)
library(here)
library (dismo)
library (rgeoboundaries)
library (ggplot2)
library (sp)
library (raster)
library (rnaturalearth)
library (rnaturalearthhires)
library(rJava)
set.seed (130824)
```

# prepare raster stack
## load countries and prepare ras mask
```{r}

# Fetch country data for specific countries
countries_1 <- ne_states(country = c("Bolivia", "Brazil", "Costa Rica", "Colombia", "Peru", "Guatemala", "Mexico"), returnclass = "sf")
countries_2 <- ne_states(country = "Argentina", returnclass = "sf")
countries_3 <- ne_states(country = c("Uruguay", "Venezuela", "Panama", "Nicaragua", "Honduras", "Belize", "El Salvador", "Trinidad and Tobago", "Suriname", "Chile", "Ecuador", "Guyana", "Paraguay"), returnclass = "sf")
 
 
admin1 <- ne_states(returnclass = "sf")
 
# Load France and filter for French Guiana
countries_4 <- admin1 %>% filter(name %in% c("Guyane française"))
 
# Extract specific territories and dependencies from admin1
france_caribbean <- admin1 %>% filter(name %in% c("Guadeloupe", "Martinique", "Saint Barthélemy", "St. Martin"))
netherlands_caribbean <- admin1 %>% filter(name %in% c("Aruba", "Curaçao", "Sint Maarten", "Bonaire", "St. Eustatius", "Saba"))
uk_caribbean <- admin1 %>% filter(admin %in% c("Anguilla", "Bermuda", "British Virgin Islands", "Cayman Islands", "Montserrat", "Turks and Caicos Islands"))
 
# US Caribbean territories
us_caribbean <- admin1 %>% filter(admin %in% c("Puerto Rico", "United States Virgin Islands
"))
 
# Caribbean countries
caribbean_countries <- admin1 %>% filter(admin %in% c("Antigua and Barbuda", "The Bahamas", "Barbados", "Cuba", "Dominica", "Dominican Republic",
                                                "Grenada", "Haiti", "Jamaica", "Saint Kitts and Nevis", "Saint Lucia", 
                                                "Saint Vincent and the Grenadines"
                                               ))
 
# Prepare data for binding by keeping only geometry and relevant columns
prepare_data <- function(df) {
  df %>% dplyr::select(geometry) %>% st_as_sf()
}
 
countries_1 <- prepare_data(countries_1)
countries_2 <- prepare_data(countries_2)
countries_3 <- prepare_data(countries_3)
countries_4 <- prepare_data(countries_4)
france_caribbean <- prepare_data(france_caribbean)
netherlands_caribbean <- prepare_data(netherlands_caribbean)
uk_caribbean <- prepare_data(uk_caribbean)
us_caribbean <- prepare_data(us_caribbean)
caribbean_countries <- prepare_data(caribbean_countries)
 
# Combine all countries and territories
countries_all <- bind_rows(
  countries_1, countries_2, countries_3, countries_4,
  france_caribbean, netherlands_caribbean, uk_caribbean, us_caribbean,
  caribbean_countries
)
 
# Check the result
plot(countries_all)
 
countries_all <- st_make_valid (countries_all)
countries_all_fixed <- countries_all |>
  st_union() %>%
  vect() %>%
  terra::fillHoles()
 

```

```{r}
ras_mask <- rast(extent = st_bbox(countries_all_fixed), 
                 resolution =0.01,
                 crs = "EPSG:4326")


ras_mask <- rasterize(countries_all_fixed, ras_mask, field = 1, background = NA)

plot (ras_mask)
ras_mask
```

## load raster data (adjust here for different years/climate projections)
###bio13_precip_wet_m
```{r}

bio13_precip_wet_m <- rast (here("outputs", "geo_proc", "chelsa_bio_13.tif"))

bio13_precip_wet_m_moll <- project(bio13_precip_wet_m,
                                ras_mask)

bio13_precip_wet_m_moll_m <- mask(bio13_precip_wet_m_moll, ras_mask)
```



###bio3_isothermality
```{r}
bio3_isothermality <-  rast (here("outputs", "geo_proc", "chelsa_bio_03.tif"))

bio3_isothermality <- bio3_isothermality/10

bio3_isothermality_moll <- project(bio3_isothermality,
                                ras_mask)

bio3_isothermality_moll_m <- mask(bio3_isothermality_moll, ras_mask)
```


##bio15_precipitation_seasonality
```{r}
bio15_precipitation_seasonality <-  rast (here("outputs", "geo_proc", "chelsa_bio_15.tif"))

bio15_precipitation_seasonality_moll <- project(bio15_precipitation_seasonality,
                                ras_mask)

bio15_precipitation_seasonality_moll_m <- mask(bio15_precipitation_seasonality_moll, ras_mask)

plot (bio15_precipitation_seasonality) # check
plot (bio15_precipitation_seasonality_moll_m) #check
```


###bio5_max_temp_warm_m
```{r}
bio5_max_temp_warm_m <- rast (here("outputs", "geo_proc", "chelsa_bio_05.tif"))
bio5_max_temp_warm_m <- bio5_max_temp_warm_m/10

bio5_max_temp_warm_m_moll <- project(bio5_max_temp_warm_m,
                                ras_mask)

bio5_max_temp_warm_m_moll_m <- mask(bio5_max_temp_warm_m_moll, ras_mask)
```

###bio17_precip_dry_q
```{r}
bio17_precip_dry_q <-  rast (here("outputs", "geo_proc", "chelsa_bio_17.tif"))

bio17_precip_dry_q_moll <- project(bio17_precip_dry_q,
                                ras_mask)

bio17_precip_dry_q_moll_m <- mask(bio17_precip_dry_q_moll, ras_mask)
```

###bio6_min_temp_cold_m
```{r}

bio6_min_temp_cold_m <-  rast (here("outputs", "geo_proc", "chelsa_bio_06.tif"))
bio6_min_temp_cold_m <- bio6_min_temp_cold_m/10

bio6_min_temp_cold_m_moll <- project(bio6_min_temp_cold_m,
                                ras_mask)

bio6_min_temp_cold_m_moll_m <- mask(bio6_min_temp_cold_m_moll, ras_mask)
```



###evergreen_broad_lc

```{r}
evergreen_broad_lc <- rast (here("data_raw", "geo_raw", "consensus_full_class_2.tif"))

evergreen_broad_lc_moll <- project(evergreen_broad_lc,
                                ras_mask)

evergreen_broad_lc_moll_m <- mask(evergreen_broad_lc_moll, ras_mask)
```


###cultivated_lc 

```{r}
cultivated_lc <- rast (here("data_raw", "geo_raw", "consensus_full_class_7.tif"))

cultivated_lc_moll <- project(cultivated_lc,
                                ras_mask)

cultivated_lc_moll_m <- mask(cultivated_lc_moll, ras_mask)
```


###pop_dens

```{r}
pop_dens <- rast (here("data_raw", "geo_raw", "gpw_v4_population_density_rev11_2015_30_sec.tif"))

pop_dens_moll <- project(pop_dens,
                                ras_mask)

pop_dens_moll_m <- mask(pop_dens_moll, ras_mask)

```

###hfp

```{r}
hfp <- rast (here("data_raw", "geo_raw", "hf_v2geo", "hdr.adf"))
hfp

hfp_moll <- project(hfp,
                                ras_mask)
hfp_moll_m <- mask(hfp_moll, ras_mask)
```


### evi

```{r}
evi <- rast (here("data_raw", "geo_raw", "evi_world_4326", "evi.tif"))
evi_moll <- project(evi,ras_mask)
evi_moll_m <- mask(evi_moll, ras_mask)
plot (evi_moll_m)

```

###altitude
```{r}
altitude <- rast (here("data_raw", "geo_raw", "wc2.1_30s_elev.tif"))



altitude_moll <- project(altitude,
                                ras_mask)
altitude_moll_m <- mask(altitude_moll, ras_mask)
```


### deforestation 

```{r}
deforestation <- rast (here("outputs", "geo_proc", "deforestation2001_2022.tif"))

deforestation <- as.factor (deforestation)

```

```{r}
deforestation_moll <- project(deforestation,
                                ras_mask)
deforestation_moll_m <- mask(deforestation_moll, ras_mask)
```

### distance_to_cities

```{r}
distance_to_cities <- rast (here("outputs", "geo_proc", "distance_pop_count_above_5000p_crop_2020_1km_4326.tif"))
distance_to_cities_moll <- project(distance_to_cities,
                                ras_mask)
distance_to_cities_moll_m <- mask(distance_to_cities_moll, ras_mask)

```


## stack
```{r}
stack_all <- c(bio13_precip_wet_m_moll_m, bio3_isothermality_moll_m, 
               bio5_max_temp_warm_m_moll_m, bio17_precip_dry_q_moll_m,bio15_precipitation_seasonality_moll_m,
               bio6_min_temp_cold_m_moll_m, evergreen_broad_lc_moll_m,   cultivated_lc_moll_m, 
pop_dens_moll_m,  hfp_moll_m, evi_moll_m, 
altitude_moll_m, deforestation_moll_m, distance_to_cities_moll_m)



```

```{r}
names(stack_all) <-c("bio13_precip_wet_m",
"bio3_isothermality",
"bio5_max_temp_warm_m",
"bio17_precip_dry_q",
"bio15_precipitation_seasonality",
"bio6_min_temp_cold_m",
"evergreen_broad_lc",
"cultivated_lc",
"pop_dens",
"hfp",
"evi",
"altitude", "deforestation", "distance_to_cities")

plot (stack_all)
```

## write Raster for later modelling
```{r}
writeRaster(stack_all,here ("outputs", "geo_proc", "rasterXXX"))
```